Genetic signatures of evolutionary rescue by a selective sweep

### 2020-03-09 manuscript/main.tex
revised ms

\usepackage{natbib}
\bibliographystyle{style/amnatnat}

@@

\usepackage[colorlinks=true, citecolor=blue, urlcolor=blue, linkcolor = blue]{hyperref}

@@

\usepackage{color}
\definecolor{darkblue}{rgb}{0.2472, 0.24, 0.6}
\definecolor{darkred}{rgb}{0.6, 0.24, 0.442893}
\newcommand{\gc}[1]{\textcolor{orange}{#1}}
\newcommand{\mo}[1]{\textcolor{red}{#1}}

@@

One of the most useful models in population genetics is that of a selective sweep and the consequent hitch-hiking of linked neutral alleles.
While variations on this model typically assume constant population size, many instances of strong selection and rapid adaptation in nature may co-occur with complex demography.
Here we extend the hitch-hiking model to evolutionary rescue, where adaptation and demography not only co-occur but are intimately entwined. 
Our results show how this feedback between demography and evolution determines -- and restricts -- the genetic signatures of evolutionary rescue, and how these differ from the signatures of sweeps in populations of constant size.
In particular, we find rescue to harden sweeps from standing variance or new mutation (but not from migration), reduce nucleotide diversity both at the selected site and genome-wide, and increase the range of observed Tajima's $D$ values.
For a given rate of population decline, the feedback between demography and evolution makes all of these differences more dramatic under weaker selection, where bottlenecks are prolonged. 
Nevertheless, it is likely difficult to infer the co-incident timing of the sweep and bottleneck from these simple signatures, never-mind a feedback between them. 
Temporal samples spanning contemporary rescue events may offer one way forward.

@@

Here we model selective sweeps in a scenario where demography and adaptive evolution are not independent.
In particular we model an instance of evolutionary rescue (\citealp{gomulkiewicz1995does}; reviewed in \citealp{bell2017evolutionary}), where a sudden environmental change causes population decline that is reverted by a selective sweep.
Under this framework, rescue is a simultaneous demographic bottleneck and selective sweep, where each affects the other.
First, because the mean absolute fitness of the population changes with the beneficial allele's frequency, the depth and duration of the bottleneck depends on the dynamics of the selective sweep, i.e., evolution affects demography.
Second, the probability that the beneficial allele establishes depends on the rate of population decline, i.e., demography affects evolution.
Together, this feedback between demography and evolution restricts the range of dynamics that are possible, and therefore also restricts the range of genetic signatures we should expect to observe.
Our goal here is to describe the range of genetic signatures that (this model of) evolutionary rescue allows, to help elucidate how rescue may obscure inferences of past selection and demography and to identify patterns that could be used to infer rescue in nature.

Most theory on evolutionary rescue to date \citep[reviewed in][]{alexander2014evolutionary} has focused on the probability of rescue.
Recently, however, some attention has been given to the dynamics of population size \citep{orr2014population}, the probability of soft sweeps \citep{wilson2017soft}, and the genetic basis of adaptation \citep{osmond2019genetic} \textit{given rescue} in haploid or asexual populations.
Here we extend this line of thinking to three modes of rescue in diploid, sexual populations, and use coalescent theory and simulations of whole chromosomes to examine the genetic signatures at linked, neutral loci.
Our focus is on three common genetic signatures: the number of unique lineages of the beneficial allele that establish (i.e., the softness of the sweep), the pattern of nucleotide diversity around the selected site (i.e., the dip in diversity), and the pattern of Tajima's $D$ around the selected site (i.e., skews in the site-frequency spectrum).
We explore three modes of rescue, where the beneficial allele arises from either standing genetic variance, recurrent \textit{de novo} mutation, or migration.
For each mode of rescue we derive the expected forward-time dynamics and resulting genetic signatures and compare these to predictions for populations of constant size as well as to individual-based simulations.
Qualitatively, we find that rescue causes faster, harder sweeps that produce wider, deeper dips in diversity and more extreme values of Tajima's $D$.
Due to the feedback between demography and evolution, the effect of rescue on the signatures of selective sweeps, relative to the signatures in populations of constant size, becomes more pronounced as the selection coefficient, and thus the probability of rescue, gets smaller.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Methods and results}
\label{sec:methods}

Code used to derive and plot all results presented below (Python scripts and Mathematica notebook) is available at \url{https://github.com/mmosmond/rescueCoalescent.git} under an MIT licence.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Deterministic trajectories}
\label{sec:rescue_forward}

Consider a population of size $N(t)$, with a beneficial allele, $A$, at frequency $p(t)$ and an ancestral allele, $a$, at frequency $q(t)=1-p(t)$.
Assume non-overlapping generations and let $W_{AA}$, $W_{Aa}$, and $W_{aa}$ be the absolute fitness (expected number of offspring) of each genotype.
Then with random mating the expected change in allele frequency \citep[equation 5.2.13 in][]{crow1970introduction} and population size in one generation is
\begin{equation}\label{eq:deltapdeltaN}
\begin{aligned}
\Delta p(t) &\equiv p(t+1) - p(t) = p(t) q(t) \frac{  (W_{AA} - W_{Aa} )p(t) +  (W_{Aa} - W_{aa})q(t)}{\overline{W}(t)}  \\
\Delta N(t) &\equiv N(t+1) - N(t) = (\overline{W}(t) - 1) N(t),
\end{aligned}
\end{equation}
with $\overline{W}(t)$ the population mean fitness.

Here we are interested in the scenario where a population composed of primarily $aa$ genotypes is declining at some rate $d$, i.e., $W_{aa} = 1-d$.
The beneficial allele, $A$, is then assumed to act multiplicatively with the fitness of the ancestral background, such that $W_{Aa} = (1-d)(1+ h s)$ and $W_{AA} = (1-d)(1+s)$.
Throughout the text we assume additivity at the selected locus, $h=1/2$, such that with weak selection the allele frequency \citep[c.f., equation 5.3.12 in][]{crow1970introduction} and population dynamics can be approximated by
\begin{equation}\label{eq:ptnt}
\begin{aligned}
p(t) &\approx \left[ 1 + \frac{1-p(0)}{p(0)} e^{-(s/2)t} \right]^{-1}\\
N(t) &\approx N(0) e^{ \left[ (1+s)(1-d)-1\right]t} \left[\frac{p(0)}{p(t)} \right]^{2(1-d)}.
\end{aligned}
\end{equation}

These equations make the feedback between evolution and demography in this model clear.
The allele frequency dynamics depend only on relative fitnesses (Equation \ref{eq:deltapdeltaN}), $W_i/\overline{W}(t)$, and thus can depend on demography only through initial allele frequency (Equation \ref{eq:ptnt}).
In contrast, demography depends on mean fitness (Equation \ref{eq:deltapdeltaN}) and thus is strongly influenced by changes in allele frequency (Equation \ref{eq:ptnt}).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Conditioning on rescue}
\label{sec:rescue_conditioning}

Above we have considered only the deterministic dynamics. 
We are, however, primarily concerned with a stochastic event -- the establishment of a rare beneficial allele.
In rescue we are particularly interested in only those instances where the beneficial allele establishes (otherwise the population is extinct), which creates a bias away from the deterministic trajectory.
As shown in \cite{smith1971use} \citep[see][for an application to evolutionary rescue]{orr2014population}, we can approximate this bias by replacing the true initial allele frequency, $p(0)$, in the deterministic predictions (Equations \ref{eq:deltapdeltaN}--\ref{eq:ptnt}) by this true value divided by the probability of establishment.
In essence, dividing initial allele frequencies by the probability of establishment implies that an allele escaping random loss when rare will initially increase much faster than expected before settling into its deterministic trajectory.

Assuming alleles do not interact (i.e., a branching process), a single copy of an allele that is expected to leave $1+\epsilon$ copies in the next generation with a variance of $\sigma^2$ has an establishment probability of \citep[][p.\ 172, see also \citealp{feller1951diffusion}, equation 5.7]{allen2010introduction}
\begin{equation}\label{eq:pest}
\rho = 1 - e^{-2 \epsilon/\sigma^2}.
\end{equation}
For example, in a Wright-Fisher population of size $N$ a rare allele with selection advantage $s$ is expected to leave $1+s$ copies, with a variance of $1+s+O(1/N)$.
Thus $\epsilon=s$ and, in a large population with weak selection, $\sigma^2 \approx 1$, such that the probability of establishment is roughly $1-e^{2s}$ \citep[][p.\ 80]{fisher1999genetical} or nearly $2s$ \citep{haldane1927mathematical}. 
In our case, $\epsilon = W_{Aa} - 1 = (1 + h s)(1-d)-1 \approx h s - d$ and $\sigma^2$ depends on the particular choice of lifecycle (e.g., see \nameref{sec:pestNe}).
When $h=1$ and $\sigma^2=1$ we recover the probability of establishment in an exponentially growing or declining population, $\rho \approx 2(s - d)$ \citep{otto1997probability}.

As mentioned above, the change in allele frequency (Equation \ref{eq:deltapdeltaN}), i.e., evolution, depends only on relative fitness and therefore is not influenced by demography.
However, now we see that conditioning on rescue will cause the initial allele frequency, and hence evolution, to depend on absolute fitness (through $\epsilon$), and hence on demography.
For example, the faster the rate of initial decline, $d$, the lower the probability of establishment (because the growth rate of the heterozygote is lower).
This causes larger effective initial allele frequencies -- and hence shorter selective sweeps -- in populations that are rescued from more precipitous declines.

The simulation details are described in full in \nameref{sec:simdetails}.
Briefly, the lifecycle described in \nameref{sec:lifecycle}, with the addition of a hard carrying capacity at $N(0)$, was simulated in \texttt{SLiM} \citep{Haller2019slim} with tree-sequence recording \citep{Haller2019tree}.
We simulated a 20 Mb segment of a chromosome, with all but one of the center loci neutral, with a per base recombination rate of $r_{bp}=2\times10^{-8}$ and per base mutation rate at neutral loci of $U=6\times10^{-9}$ (both inspired by \textit{Drosophila} estimates).
A population was considered rescued when the beneficial mutation was fixed and the population size had recovered to it's initial size, $N(0)$.
Pairwise nucleodtide diversity and Tajima's $D$ were calculated directly from the tree sequences using \texttt{msprime} \citep{kelleher2016efficient} and \texttt{tskit} \cite[][]{kelleher2018efficient}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Forward-time results}
\label{sec:fwd_results}

Below we derive the effective initial allele frequency given rescue (allowing us to predict the forward-time dynamics; see \nameref{sec:rescue_conditioning}) and the softness of the sweep given rescue, for three types of rescue: from standing genetic variation, recurrent \textit{de novo} mutation, or migration.
To do so we must first compute the probability of rescue, which we also report.
Many similar results have been previously derived: \cite{orr2008population} give the probability of rescue from standing genetic variation (haploid) or \textit{de novo} mutation (haploid and diploid); \cite{orr2014population} give the effective initial allele frequencies given rescue from standing genetic variation or \textit{de novo} mutation in a haploid model; and \cite{wilson2017soft} give the probability of a soft sweep given rescue from \textit{de novo} mutation in a haploid model.
Nevertheless, we include the diploid versions of these results below for completeness, and because they are needed to describe the genetic signatures of rescue that follow.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Rescue from standing genetic variation (SGV)}
\label{sec:rescue_forward_SGV}

We first consider rescue from genetic variation that is present at the time of the environmental change, ignoring further mutations.
To avoid complicating the presentation here we assume the initial number of beneficial alleles is given; treating the initial number as a random variable requires conditioning the distribution on a successful sweep \citep[][]{hermisson2017soft}.

Given there are initially $\kappa\ll N(0)$ copies of the beneficial allele, the number that establish is roughly binomially distributed with $\kappa$ trials and success probability $\rho$.
The probability of rescue is the probability that at least one establishes,
\begin{equation}
P_\mathrm{rescue}^\mathrm{SGV} = 1-(1-\rho)^\kappa.
\end{equation}
Conditioning on rescue, it is therefore as if the true initial allele frequency, $\kappa/(2N(0))$, is inflated by a factor, $1/P_\mathrm{rescue}^\mathrm{SGV}$, i.e., the beneficial allele quickly jumps to frequency \citep[c.f., equation S1.4 in][]{orr2014population}
\begin{equation}\label{eq:p0sgv}
p_{0|\mathrm{rescue}}^{\mathrm{SGV}} = \frac{\kappa}{2N(0)} \frac{1}{P_{\mathrm{rescue}}^{\mathrm{SGV}}},
\end{equation}
after which it follows its deterministic trajectory.
This same conditioning applies in a population of constant size (i.e., $d=0$).
As the decline rate, $d$, increases, the probability a copy establishes, $\rho$, and hence the probability of rescue, $P_\mathrm{rescue}^\mathrm{SGV}$, declines, making the conditioning stronger.
This causes selective sweeps to get started faster as $d$ increases, implying that, for a given $s$, rescue sweeps are faster than those in populations of constant size.

Figure \ref{fig:rescueSGV_dynamics} compares our numerical (Equation \ref{eq:deltapdeltaN}) and analytical (Equations \ref{eq:ptnt}) approximations against individual-based simulations.
This shows that the numerical predictions do a reasonably good job in capturing the median simulation dynamics (we use the median as taking the mean of the non-linear trajectories distorts their shape), especially with larger selection coefficients and more initial copies of the beneficial allele.
For smaller selection coefficients and fewer initial copies of the beneficial allele we overestimate allele frequency and population size (e.g., panel D).
The reasons for this are two-fold.
On the one hand we tend to underestimate the probability of establishment with small selection coefficients; beneficial alleles with smaller selection coefficients can exist at low numbers -- before securing establishment -- for a longer period of time \citep[][e.g., with $s=0.13$ and $d=0.05$ we have $W_{Aa} \approx 1.01$, so that such an allele can persist for up to $\sim(W_{Aa}-1)^{-1} \approx 100$ generations before establishing or going extinct, \citealp{desai2007beneficial}]{maruyama1974note}.
The longer it takes a beneficial allele to establish the fewer copies of the ancestral allele that exist when it does (in the case of rescue, $d>0$), increasing the chance the beneficial allele will experience some selection as a homozygote while establishing (where it has roughly twice the probability of survival when $d$ is small given $h=1/2$). 
This is true even when we start with only a single copy of the beneficial allele (as opposed to a model with constant population size, $d=0$, where the ancestral allele only decreases in frequency if the beneficial allele increases in number), although this effect is strengthened when there are initially more copies as some copies may reach substantial frequencies in the meantime, causing even greater beneficial allele frequencies during establishment.
On the other hand, conditioning on rescue has a larger effect (on effective initial allele frequencies) when rescue is unlikely.
This amplifies our underestimates of the probability of establishment discussed above when there are initially fewer copies of the beneficial allele.
The analytical approximations, assuming weak selection, provide yet larger overestimates of allele frequency and population size (due to the relatively large selection coefficients used, as expected in populations maladapted enough to require evolutionary rescue).

>>>>

\usepackage{natbib}
\bibliographystyle{amnatnat}

@@

\usepackage[colorlinks=true, citecolor=blue, urlcolor=blue, linkcolor=blue]{hyperref}

@@

\usepackage{color}
\definecolor{darkblue}{rgb}{0.2472, 0.24, 0.6}
\definecolor{darkred}{rgb}{0.6, 0.24, 0.442893}
\definecolor{darkyellow}{rgb}{0.6, 0.547014, 0.24}

@@

One of the most useful models in population genetics is that of a selective sweep and the consequent hitch-hiking of linked neutral alleles.
While variations on this model typically assume constant population size, many instances of strong selection and rapid adaptation in nature may co-occur with complex demography.
Here we extend the hitch-hiking model to evolutionary rescue, where adaptation and demography not only co-occur but are intimately entwined. 
Our results show how this feedback between demography and evolution determines -- and restricts -- the genetic signatures of evolutionary rescue, and how these differ from the signatures of sweeps in populations of constant size.
In particular, we find rescue to harden sweeps from standing variance or new mutation (but not from migration), reduce nucleotide diversity both at the selected site and genome-wide, and increase the range of observed Tajima's $D$ values.
For a given initial rate of population decline, the feedback between demography and evolution makes all of these differences more dramatic under weaker selection, where bottlenecks are prolonged. 
Nevertheless, it is likely difficult to infer the co-incident timing of the sweep and bottleneck from these simple signatures, never-mind a feedback between them. 
Temporal samples spanning contemporary rescue events may offer one way forward.

@@

Here we model selective sweeps in a scenario where demography and adaptive evolution are not independent.
In particular we model an instance of evolutionary rescue (\citealp{gomulkiewicz1995does}; reviewed in \citealp{bell2017evolutionary}), where a sudden environmental change causes population decline that is reverted by a selective sweep.
Under this framework, rescue is a simultaneous demographic bottleneck and selective sweep, where each affects the other.
First, because the mean absolute fitness of the population changes with the beneficial allele's frequency, the depth and duration of the bottleneck depends on the dynamics of the selective sweep, i.e., evolution affects demography.
Second, the probability that the beneficial allele establishes depends on its growth rate and thus the rate of population decline, i.e., demography affects evolution.
Together, this feedback between demography and evolution restricts the range of dynamics that are possible, and therefore also restricts the range of genetic signatures we should expect to observe.
Our goal here is to describe the range of genetic signatures that (this model of) evolutionary rescue allows, to help elucidate how rescue may obscure inferences of past selection and demography and to identify patterns that could be used to infer rescue in nature.

Most theory on evolutionary rescue to date \citep[reviewed in][]{alexander2014evolutionary} has focused on the probability of rescue.
Recently, however, some attention has been given to the dynamics of population size \citep{orr2014population}, the probability of soft sweeps \citep{wilson2017soft}, and the genetic basis of adaptation \citep{osmond2020genetic} \textit{given} rescue in haploid or asexual populations.
Here we extend this line of thinking to three modes of rescue in diploid, sexual populations, and use coalescent theory and simulations of whole chromosomes to examine the genetic signatures at linked, neutral loci.
Our focus is on three common genetic signatures: the number of unique lineages of the beneficial allele that establish (i.e., the softness of the sweep), the pattern of nucleotide diversity around the selected site (i.e., the dip in diversity), and the pattern of Tajima's $D$ around the selected site (i.e., skews in the site-frequency spectrum).
We explore three modes of rescue, where the beneficial allele arises from either standing genetic variance, recurrent \textit{de novo} mutation, or migration, and compare to selective sweeps arising in populations of constant size.
Qualitatively, we find that for a given selection coefficient rescue causes faster, harder sweeps that produce wider, deeper dips in diversity and more extreme values of Tajima's $D$.
Due to the feedback between demography and evolution, the effect of rescue on the signatures of selective sweeps becomes more pronounced as the selection coefficient, and thus the probability of rescue, gets smaller.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Methods}
\label{sec:methods}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Population dynamics}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Deterministic trajectories}
\label{sec:rescue_forward}

Consider a population of size $N(t)$ with a beneficial allele, $A$, at frequency $p(t)$ and an ancestral allele, $a$, at frequency $q(t)=1-p(t)$.
Assume non-overlapping generations and let $W_{AA}$, $W_{Aa}$, and $W_{aa}$ be the absolute fitness (expected number of offspring) of each genotype.
Then with random mating the expected change in allele frequency \citep[equation 5.2.13 in][]{crow1970introduction} and population size in one generation is
\begin{equation}\label{eq:deltapdeltaN}
\begin{aligned}
\Delta p(t) &\equiv p(t+1) - p(t) = p(t) q(t) \frac{  (W_{AA} - W_{Aa} )p(t) +  (W_{Aa} - W_{aa})q(t)}{\overline{W}(t)}  \\
\Delta N(t) &\equiv N(t+1) - N(t) = (\overline{W}(t) - 1) N(t),
\end{aligned}
\end{equation}
with $\overline{W}(t)$ the population mean fitness.

Here we are interested in the scenario where a population composed of primarily $aa$ genotypes is declining at some rate $d$, i.e., $W_{aa} = 1-d$.
The beneficial allele, $A$, is then assumed to act multiplicatively with the fitness of the ancestral background, such that $W_{Aa} = (1-d)(1+ h s)$ and $W_{AA} = (1-d)(1+s)$.
Throughout the text we assume additivity at the selected locus, $h=1/2$, such that the allele frequency \citep[c.f., equation 5.3.12 in][]{crow1970introduction} and population dynamics can be approximated by
\begin{equation}\label{eq:ptnt}
\begin{aligned}
p(t) &\approx \left( 1 + \frac{1-p(0)}{p(0)} e^{-s t/2} \right)^{-1}\\
N(t) &\approx N(0) e^{ \left[ (1+s)(1-d)-1\right]t} \left(\frac{p(0)}{p(t)} \right)^{2(1-d)}.
\end{aligned}
\end{equation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Effective initial allele frequency}
\label{sec:initial_frequency}

Equations \ref{eq:deltapdeltaN} and \ref{eq:ptnt} consider only the deterministic dynamics. 
We are, however, primarily concerned with a stochastic event -- the establishment of a (potentially rare) beneficial allele.
In rescue we are particularly interested in only those instances where the beneficial allele establishes (otherwise the population is extinct), which creates a bias away from the deterministic trajectory.
As shown in \cite{smith1971use} \citep[see][for an application to evolutionary rescue]{orr2014population}, we can approximate this bias by assuming that a single copy of an allele that establishes with probability $P_\mathrm{est}$ instantaneously increases from a frequency of $1/(2N)$ to $1/(2NP_\mathrm{est})$.
In essence, dividing the initial allele frequency by the probability of establishment implies that an allele escaping random loss when rare will quickly increase to a larger frequency than expected before settling into its deterministic trajectory.
As we will see below, we can combine this logic with the waiting time distribution for an establishing mutation to arrive to derive an effective initial allele frequency, $p_0$.
Using the deterministic equations (Equation \ref{eq:deltapdeltaN} or \ref{eq:ptnt}) with $p(0) = p_0$ then approximates the forward-time trajectory of an allele that is destined to fixation.

We derive the probability of establishment, $P_\mathrm{est}$, assuming alleles do not interact (i.e., a branching process).
Under this assumption a single copy of an allele that is expected to leave $1+\epsilon$ copies in the next generation with a variance of $v$ has an establishment probability of \citep[][p.\ 172, see also \citealp{feller1951diffusion}, equation 5.7, for a derivation from a diffusion approximation]{allen2010introduction}
\begin{equation}\label{eq:pest}
P_\mathrm{est} = 1 - e^{-2 \epsilon/v}.
\end{equation}
For example, in a Wright-Fisher population of size $N$ a rare allele with selection advantage $s$ is expected to leave $1+s$ copies, with a variance of $1+s+O(1/N)$.
Thus $\epsilon=s$ and, in a large population with weak selection, $v \approx 1$, such that the probability of establishment is roughly $1-e^{2s}$ \citep[][p.\ 80]{fisher1999genetical} or nearly $2s$ \citep{haldane1927mathematical}. 
In our case, in a declining population ($d>0$) we have $\epsilon = W_{Aa} - 1 = (1 + s h)(1-d)-1 \approx s h - d$ and $v$ depends on the particular choice of lifecycle (e.g., see \nameref{sec:pestNe}).
When $h=1$ and $\sigma^2=1$, we recover the probability of establishment of a weakly beneficial mutation in an exponentially growing or declining haploid population with a Poisson offspring distribution, $P_\mathrm{est} \approx 2(s - d)$ \citep{otto1997probability}.
In a population at carrying capacity (e.g., with $d\geq0$) the growth rate is instead $\epsilon = W_{Aa}/W_{aa} - 1 = (1 + s h)-1 = s h$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Effective final allele frequency}
\label{sec:final_frequency}

Above we have stitched together the initial stochastic phase of the allele frequency and population dynamics with the deterministic phase by using $p(0)=p_0$. 
We next incorporate the final phase, fixation, which is also stochastic.
Incorporating this final phase is important here \citep[as opposed to studies that predict only the forward-time dynamics; e.g.,][]{orr2014population} because in looking at genetic signatures we are primarily concerned with the dynamics looking backwards in time from the point of fixation, and this third phase determines how fast fixation occurs.
As shown in \cite{martin2015simple}, we can treat the backward-in-time dynamics from fixation as we did the forward-in-time dynamics at establishment. 
In particular, the probability the heterozygote establishes (backward in time) in a population of mutant homozygotes is $P_\mathrm{est}$ (Equation \ref{eq:pest}) with the growth rate of a rare heterozygote in this population, $\epsilon$, replaced by $-\epsilon$ (let's call this probability $P_\mathrm{est}^-$).
For instance, if the fitness of the heterozygote is $W_{Aa} = (1-d)(1+ s h)$ and the fitness of the mutant homozygote is $W_{AA} = (1-d)(1+s)$ and the population is at carrying capacity then $-\epsilon = 1 - W_{Aa}/W_{AA}=1 - (1+ s h)/(1+ s) \approx s(1-h)$.
This does not depend on demography, $d$, in contrast to the growth rate of the heterozygote at establishment (see \nameref{sec:initial_frequency}).
Note that the variance in the number of offspring, $v$, may also be different at the time of fixation than it was at the time of establishment (see \nameref{sec:pestNe}).
Given the heterozygote establishes it will quickly increase to $1/P_\mathrm{est}^-$ copies, implying that fixation effectively occurs when the deterministic trajectory reaches a frequency of $p_f = 1-1/(2 N P_\mathrm{est}^-)$.
Below we will explore scenarios where fixation is expected to occur while the population is at carrying capacity. 
Thus in all cases we have the same effective final allele frequency,
\begin{equation}\label{eq:pf}
p_f = 1 - \frac{1}{2 N(0) P_\mathrm{est}},
\end{equation}
where $P_\mathrm{est}$ is given by Equation \ref{eq:pest} with $\epsilon = 1 - (1 + s h)/(1 +  s)$ and $v$ as given in \nameref{sec:pestNe}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Time to fixation}
\label{sec:fixation_time}

We can estimate the time to fixation in the additive model ($h=1/2$) as the time it takes the deterministic approximation (Equation \ref{eq:ptnt}) to go from the effective initial frequency, $p_0$, to the effective final frequency, $p_f$, which is \citep[equation 5.3.13 in ][]{crow1970introduction}
\begin{equation}\label{eq:tfix}
t_f = \frac{2}{s} \log\left[ \frac{(1-p_0)p_f}{p_0(1-p_f)}\right].
\end{equation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Backward-time dynamics}
\label{sec:backward_dynamics}

Setting $t = t_f - \tau$ in Equation \ref{eq:ptnt} gives an approximation of the dynamics of allele frequency and population size backward in time,
\begin{equation}\begin{aligned}\label{eq:backwardsptnt}
p^\prime(\tau) &\approx \left(1 + \frac{1 - p_f}{p_f} e^{s \tau/2} \right)^{-1}\\
N^\prime(\tau) &\approx N(0) e^{-d (t_f-\tau)} \left(\frac{1-p_0}{1-p^\prime(\tau)}\right)^{2(1-d)},
\end{aligned}\end{equation}
from the point of fixation, $\tau=0$, to the time of establishment, $\tau=t_f$.
Note that the backward-time allele frequency dynamics do not depend on the initial frequency or $d$ except in determining the maximum value of $\tau$.
Thus the backward-time allele frequency dynamics are expected to follow the same trajectory in rescue as in populations of constant size with the same carrying capacity (such that the final frequency, $p_f$, is the same).
Meanwhile, the backward-time population size dynamics depend heavily on the initial frequency as the timing of the sweep determines the depth and duration of the bottleneck.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{The structured coalescent}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Event rates}
\label{sec:rescue_backward}

To explore the genetic patterns created by evolutionary rescue we next consider a random sample of chromosomes at the time the beneficial allele fixes.
Focusing on a neutral locus that is recombination distance $r$ from the selected site, we are interested in calculating the rate of coalescence, the rates of recombination and mutation off the selected background, and the rate of migration out of the population.
If our sample of alleles has $k$ distinct ancestors on the selected background $\tau$ generations before fixation, these rates are approximately (table 1 in \citealp{hudson1988coalescent}; equation 16 in \citealp{pennings2006soft2}; see \nameref{sec:derive_coalescent} for derivations)
\begin{equation}\label{eq:coal_rates}
\begin{aligned}
p_{\mathrm{coal}}(k,\tau) &= \binom{k}{2} \frac{1}{2N_e'(\tau) p^\prime(\tau)}\\
p_{\mathrm{rec}}(k,\tau) &= k r [1 - p^\prime(\tau)]\\
p_{\mathrm{mut}}(k,\tau) &= k \frac{u [1 - p^\prime(\tau)]}{p^\prime(\tau)}\\
p_{\mathrm{mig}}(k,\tau) &= k \frac{m}{2N^\prime(\tau)p^\prime(\tau)},
\end{aligned}
\end{equation} 
where $N_e'(\tau)$ is the effective population size $\tau$ generations before fixation.

With slow changes in population size, $N(t-1)\approx N(t)$ and the mean number of gametes contributed to the next generation by each diploid individual in the current generation is $2$.
The inbreeding and variance effective population size, $N_e(t)$, is then roughly $(4N(t) - 2)/(\sigma^2+2)$ \citep[equation 7.6.4.3 in][]{crow1970introduction}, where $\sigma^2$ is the variance in the per capita number of gametes contributed to the next generation.
Therefore, in a large population, $N(t)\gg1$, the ratio of the effective size to the census size is roughly $N_e(t)/N(t) \approx 4/(2+\sigma^2)$, where $\sigma^2$ depends on the particular lifecycle (e.g., see \nameref{sec:pestNe}).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Event timing}

We now use the instantaneous event rates (Equation \ref{eq:coal_rates}) to calculate the probability that the most recent event is $\tau$ generations before fixation and is either coalescence, recombination, mutation, or migration. 
Letting $i,j\in\{\mathrm{coal, rec, mut, mig}\}$, the probability that $i$ is the most recent event, and occurs $\tau$ generations before fixation, is \citep[c.f., equation 6 in][]{pennings2006soft}
\begin{equation}\label{eq:Pikt}
\begin{aligned}
P_i(k,\tau) &= p_{i}(k,\tau) \left( \prod_{j\neq i} [1 - p_{j}(k,\tau)] \right) \prod_{l=0}^{\tau-1} \left( \prod_{j} [1 - p_{j}(k,l)] \right)\\
&\approx p_{i}(k,\tau) \exp \left( -\sum_{j} \int_{0}^{\tau} p_{j}(k,l) \mathrm{d}l \right),
\end{aligned}
\end{equation}
i.e., the waiting time for an inhomogeneous exponential random variable.
The approximation assumes the $p_i(k,\tau)$ are small enough such that at most one event happens each generation, with small probability, and the changes in the $p_i(k,\tau)$ from one generation to the next are small enough that we can approximate a sum across $\tau$ with an integral.
As a technical aside, to speed computation we analytically solve the integrals and evaluate Equation \ref{eq:Pikt} under the assumption of weak selection and exponential population growth (Equation \ref{eq:ptnt}).
